---
title: "Simple Things in Python and R"
output: md_document
permalink: /beginnings/
questions:
  - "How do I print things?"
  - "What are R's basic data types?"
  - "How do I find out what type something is?"
  - "How do I name variables in R?"
  - "How do I create and index lists in R?"
  - "How do I write conditionals and loops in R?"
  - "What should I use instead of loops in R?"
  - "How do ranges in R differ from ranges in Python?"
  - "What special values does R use to represent things that aren't there?"
  - "How do I create functions in R?"
objectives:
  - "Name and describe R's atomic data types and create objects of those types."
  - "Explain what 'scalar' values actually are in R."
  - "Identify correct and incorrect variable names in R."
  - "Create vectors in R and index them to select single values, ranges of values, and selected values."
  - "Create `for` loops and `if`/`else` statements in R."
  - "Explain why vectors cannot be used directly in conditional expressions and correctly use `all` and `any` to combine their values."
  - "Define functions taking a fixed number of named arguments and/or a variable number of arguments."
  - "Explain the difference between `NA` and `NULL` and correctly use tests for each."
  - "Explain what vectorization is and create vectorized equivalents of unnested loops containing simple conditional tests."
keypoints:
  - "Use `print(expression)` to print the value of a single expression."
  - "Variable names may include letters, digits, `.`, and `_`, but `.` should be avoided, as it sometimes has special meaning."
  - "R's atomic data types include logical, integer, double (also called numeric), and character."
  - "R stores collections in homogeneous vectors of atomic types, or in heterogeneous lists."
  - "'Scalars' in R are actually vectors of length 1."
  - "Vectors and lists are created using the function `c(...)`."
  - "Vector indices from 1 to length(vector) select single elements."
  - "Negative indices to vectors deselect elements from the result."
  - "The index 0 on its own selects no elements, creating a vector or list of length 0."
  - "The expression `low:high` creates the vector of integers from `low` to `high` inclusive."
  - "Subscripting a vector with a vector of numbers selects the elements at those locations (possibly with repeats)."
  - "Subscripting a vector with a vector of logicals selects elements where the indexing vector is `TRUE`."
  - "Values from short vectors (such as 'scalars') are repeated to match the lengths of longer vectors."
  - "The special value `NA` represents missing values, and (almost all) operations involving `NA` produce `NA`."
  - "The special values `NULL` represents a nonexistent vector, which is not the same as a vector of length 0."
  - "Use `for (loop_variable in collection){ ...body... }` to create a loop."
  - "Use `if (expression) { ...body... } else if (expression) { ...body... } else { ...body... }` to create conditionals."
  - "Expression conditions must have length 1; use `any(...)` and `all(...)` to collapse logical vectors to single values."
  - "Most operators and functions in R work on corresponding elements of vectors, and should be used in preference to loops."
  - "Use `ifelse(vector_condition, values_if_true, values_if_false)` in place of conditionals inside loops."
  - "Use `function(...arguments...) { ...body... }` to create a function."
  - "Use variable <- function(...arguments...) { ...body... }` to create a function and give it a name."
  - "The body of a function can be a single expression or a block in curly braces."
  - "The last expression evaluated in a function is returned as its result."
  - "Use `return(expression)` to return a result early from a function."
  - "Use `sapply(vector, function)` to apply a function to each value in a vector in turn, returning a vector of results."
  - "Use `sapply(vector, function(x){ ...body... })` to perform simple operations on each element of a vector."
---

Let's try some simple things at an interactive prompt.

## How do I greet the world?

Let us begin as is traditional,
using pink to show Python code and green to show R:

```{python python-hello}
"Hello, world!"
```

```{r r-hello}
print("Hello, world!")
```

A sense of weariness settles over us immediately.
Python prints exactly what we have asked for,
but what is that `[1]` in R's output?
Is it perhaps something akin to a line number?
Let's take a closer look by simply evaluating an expression without bothering with `print`:

```{r r-quotes}
'This is in single quotes.'
"This is in double quotes."
```

It appears that R uses double quotes to display strings even when we give it a single-quoted string
(which is no worse than Python using single quotes when we've given it doubles).
The bad news is that the mysterious `[1]` doesn't appear to be a line number.
Let's ignore it for now and do a little more exploring.

## How do I add numbers?

In Python,
I add numbers using `+`.
(Note that, as above, we use `print` to display the output instead of simply typing in the expression.
This is a feature of the current version of RStudio.)

```{python python-addition}
print(1 + 2 + 3)
```

I can check the [type](../glossary/#type) of the result using the built-in `type` function,
which in this case tells me that my `6` is an [integer](../glossary/#integer):

```{python python-type}
print(type(6))
```

What does R do?

```{r r-addition}
1 + 2 + 3
```
```{r r-typeof}
typeof(6)
```

Hm.
Calling the type-checking function `typeof` rather than `type` isn't frightening,
and having it return the name of a type rather than a class can also be excused,
but it does seem odd for integer addition to produce a double-precision [floating-point](../glossary/#floating-point) result
(which is what the type `"double"` means).
Let's try an experiment:

```{r}
typeof(6)
```

Ah ha.
By default,
R represents numbers as floating-point values,
even if they look like integers when written.
We can force a [literal value](../glossary/#literal) to be an integer by putting an upper-case L after it:

```{r}
typeof(6L)
```

Arithmetic on integers produces integers:

```{r}
1L + 2L + 3L
```
```{r}
typeof(1L + 2L + 3L)
```

and if we want to convert a "normal" (floating-point) number to an integer
we can do so:

```{r}
as.integer(6)
```
```{r}
typeof(as.integer(6))
```

But wait:
what is that dot doing in that function's name?
Is there an object called `as` with a [method](../glossary/#method) called `integer`?
No.
In R,
`.` is just another character.
Like the underscore `_`,
it is used to make names more readable,
but it has no special meaning.

## How do I store many numbers together?

The Elder Gods do not bother to learn most of our names because there are so many of us and we are so...ephemeral.
Similarly, we only give proper names to a handful of values in our programs;
we lump the rest together into lists and matrices and more esoteric structure so that
we too can create, manipulate, and dispose of multitudes with a single dark command.

In Python,
we create a [list](../glossary/#list) using square brackets,
and [assign](../glossary/#assignment) that list to a [variable](../glossary/#variable) using `=`.
If the variable does not exist, it is created:

```{python}
primes = [3, 5, 7, 11]
```

Since assignment is a statement rather than an expression,
it has no result,
so Python does not display anything when this command is run.

The equivalent operation in R uses a function called `c`,
which stands for "column" and which creates a [vector](../glossary/#vector):

```{r}
primes <- c(3, 5, 7, 11)
```

Assignment is done using a left-pointing arrow `<-`.
Like Python,
R does not display a value after an assignment statement.

Now that we can create vector in R,
we can explain that errant `[1]` in our previous examples.
To begin with,
let's have a look at the lengths of various things in Python:

```{python}
print(len(primes))
```

```{python error=TRUE}
print(len(4))
```

Fair enough:
the length of a list is the number of elements it contains,
and since a [scalar](../glossary/#scalar) value like the integer 4 doesn't contain elements,
we are committing a minor sin by asking for its length.

Now,
what of R and its vectors?

```{r}
length(primes)
```

Good.

```{r}
length(4)
```

That's unexpected.
Let's have a closer look:

```{r}
typeof(primes)
```

That's also unexpected,
but all becomes clear once we realize that *there are no scalars in R*.
`4` is not a single lonely integer,
but rather a vector of length one containing the value 4.
When we display its value,
the `[1]` is the index of its first value.
We can prove this by creating and displaying a much longer vector:

```{r}
longer <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
longer
```

In order to help us find out way in our data,
R automatically breaks long lines
and displays the starting index of each line.
These indices also show us that R counts from 1 as humans do,
rather than from zero.
(There are a great many myths about why programming languages do the latter.
If you dare know the truth,
[Mike Hoye has it for you][hoye-zero].)

## How do I index a ~~list~~ vector?

Python's rules for [indexing](../glossary/#indexing) are simple once you understand them
(a statement which is also true of quantum mechanics).
To avoid confusing indices with values,
let's create a list of color names and index that:

```{python}
colors = ["eburnean", "glaucous", "wenge"]
print(colors[0])
```
```{python}
print(colors[2])
```
```{python error=TRUE}
colors[3]
```
```{python}
print(colors[-1])
```

Indexing the equivalent vector in R with the indices 1 to 3 produces unsurprising results:

```{r}
colors <- c("eburnean", "glaucous", "wenge")
colors[1]
```
```{r}
colors[3]
```

What happens if we go off the end?

```{r}
colors[4]
```

Our minds cannot grasp reality without shattering into a million gibbering pieces.
In statistics,
this lack of omniscience is often reflected by gaps in data.
R uses the special value [`NA`](../glossary/#NA) (short for "not available") to represent these gaps,
and tries to shield us from the full horror of the unknowable by returning `NA`
when we ask for knowledge that does not exist.

R does more than this---much more.
In Python,
a negative index counts backward from the end of a list.
In R,
we use a negative index to indicate a value that we don't want:

```{r}
colors[-1]
```

But wait.
If every value in R is a vector,
then when we use 1 or -1 as an index,
we're actually using a vector to index another one.
What happens if the index itself contains more than one value?

```{r error=TRUE}
colors[1, 2]
```

All right, that didn't work.
What if we make the vector subscript explicit using `c`?

```{r}
colors[c(3, 1, 2)]
```
```{r}
colors[c(1, 1, 1)]
```
```{r}
colors[c(-1, -2)]
```
```{r error=TRUE}
colors[c(1, -1)]
```

This looks promising.
If we provide a vector of positive indices,
we get the elements they identify in their order,
with elements duplicated if an index was used multiple times.
The same thing happens if we use negative indices,
and if we try to mix them,
R heaves a heavy sigh and chides us for our mistake,
presumably because an index like `c(1, -1)` is ambiguous.

What of zero?

```{r}
colors[0]
```

In order to understand this rather cryptic response,
we can try calling the function `character` ourselves
with a positive argument:

```{r}
character(3)
```

Ah---it appears that `character` constructs a vector of character strings of the specified length
and fills it with empty strings.
The expression `character(0)` presumably therefore means
"an [empty vector](../glossary/#empty-vector) of type character".
From this,
we conclude that the index 0 doesn't correspond to any elements,
so R gives us back something of the right type but with no content.
As a check,
let's try indexing with 0 and 1 together:

```{r}
colors[c(0, 1)]
```

This unfortunate behavior is fertile ground for breeding bugs.
Guard against it as you would against the Hounds of Tindalos.

## How do I choose and repeat things?

We cherish the illusion of free will so much that we embed a pretense of it in our machines
in the form of [conditional statements](../glossary/#conditional).
We then instruct those same machines to make the same decisions over and over,
often for no discernible purpose.
For example,
we could write the following [for loop](../glossary/#for-loop) in Python:

```{python}
values = [-1, 0, 1]
for v in values:
    if v < 0:
        sign = -1
    elif v == 0:
        sign = 0
    else:
        sign = 1
    print("The sign of", v, "is", sign)
```

and then create something equivalent in R:

```{r}
values <- c(-1, 0, 1)
for (v in values) {
  if (v < 0) {
    sign <- -1
  }
  else if (v == 0) {
    sign <- 0
  }
  else {
    sign <- 1
  }
  print(paste("The sign of", v, "is", sign))
}
```

There are a few things to note here:

1.  The parentheses in the loop header are required:
    we cannot simply write `for v in values`.
2.  The curly braces around the [body](../glossary/#loop-body) of the loop
    and the [conditional branches](../glossary/#conditional-branch) are optional,
    but should always be there to help readability.
3.  `paste` converts its arguments to strings and then concatenates those,
    placing a single space between each unless instructed to do otherwise.
    `print` then prints the resulting (single) string.
    The functions `cat` and `message` can also be used for text output.
4.  By calling our temporary variable `sign`
    we risk [collision](../glossary/#name-collision) with the rather useful built-in R function of the same name.

Explicit loops in R programs are generally regarded with the same sort of suspicion
as hearty claims by over-confident graduate students that
the text they are about to read aloud is purely metaphorical.
If we wish to get the signs of some values in R,
we can more idiomatically do this:

```{r}
result <- sign(c(-1, 0, 1))
```

Vectors being ubiquitous in R,
almost every function will operate on multiple values.
And as sequences such as (-1, 0, 1) are also very common,
R provides a notation for expressing them:

```{r}
result <- sign(-1:1)
```

Where Python only allows [ranges](../glossary/#range-expression) such as `-1:1` to be used as indices,
R considers them first-class entities.
Their most common use is as indices to vectors:

```{r}
colors <- c("eburnean", "glaucous", "squamous", "wenge")
colors[1:3]
```

We can similarly subtract a range of colors by index:

```{r}
colors[-1:-3]
```

R does not allow tripartite expressions of the form `start:end:stride`;
for that,
we must use the `seq` function:

```{r}
seq(1, 10, 3)
```

This example also shows that ranges in R are inclusive at both ends,
i.e.,
they run up to *and including* the upper bound.
As is traditional among programming language advocates,
we claim that this is more natural,
and then cite as proof some supportive anecdote such as,
"Most people do not interpret the expression 'from one to five'
to mean 'one, two, three, or four'."

What we *cannot* do is use vectors directly as conditions in `if` statements:

```{r}
numbers <- c(0, 1, 2)
if (numbers) {
  print("This should not have worked.")
}
```

Instead,
we must collapse them into a single [logical value](../glossary/#Boolean-value)
(often called a Boolean in honor of a logician who did not go mad contemplating the nature of truth):

```{r}
numbers <- c(0, 1, 2)
if (all(numbers >= 0)) {
  print("This, on the other hand, should work.")
}
```

When evaluating the expression `numbers >= 0`,
R automatically replicates or [recycles](../glossary/#recycling) the shorter vector
(in this case, the vector of length 1 containing the value 0)
and then performs element-by-element comparison to construct a vector of logical values:

```{r}
numbers >= 0
```

The function `all` then checks if they are all `TRUE`.
We could use a corresponding function `any` to check if at least one was `TRUE`;
these two functions therefore correspond to "and" and "or" across the whole vector.

Logical vectors have a special place in the R universe,
just as human beings do not in the larger one.
If a vector is [indexed with a logical vector](../glossary/#logical-indexing),
the result is a vector containing only the values at locations where the index vector is `TRUE`.
This is easier to show than to describe:

```{r}
both <- -2:2         # c(-2, -1, 0, 1, 2)
mask <- both > 0     # c(FALSE, FALSE, FALSE, TRUE, TRUE)
result <- both[mask] # c(1, 2)
result
```

We can of course make the code shorter by eliminating `mask`:

```{r}
result <- both[both > 0]
result
```

## How do I create and call functions?

As we have already seen,
we call [functions](../glossary/#function) in R much as we do in Python:

```{r}
max(1, 3, 5) + min(1, 3, 5)
```

We can define a new function using the `function` keyword.
Doing so does not name the function;
to accomplish that,
we assign the newly-created function to a variable:

```{r}
swap <- function(pair) {
  c(pair[2], pair[1])
}
swap(c("left", "right"))
```

As this example shows,
the result of a function is the value of the last expression evaluated within it.
A function can return a value earlier using the `return` function;
we can use `return` for the final value as well,
but most R programmers do not.

```{r}
swap <- function(pair) {
  if (length(pair) != 2) {
    return(NULL) # This is very bad practice.
  }
  c(pair[2], pair[1])
}
swap(c("one"))
```

```{r}
swap(c("left", "right"))
```

We pause now to answer some questions that may have occurred to attentive readers.
First,
the special value [`NULL`](../glossary/#null) represents the absence of a vector.
It is not the same as a vector of zero length,
though testing that statement produces a rather odd result:

```{r}
NULL == integer(0)
```

To test for nullity,
use the function `is.null`:

```{r}
is.null(NULL)
```

Second,
and more importantly,
returning `NULL` when our function's inputs are invalid is an express route to madness,
as doing so means that `swap` can fail without telling us that it has done so.
Consider:

```{r}
NULL[1]  # Try to access an element of the vector that does not exist.
```

```{r}
values <- 5:10          # More than two values.
result <- swap(values)  # Attempting to swap the values produces NULL.
result[1]               # But we can operate on the result without error.
```

The next section will explore what we ought to do instead.
Before then,
however,
we can offer some slight comfort.
If the number of arguments given to a function is not the number expected,
R will complain:

```{r error=TRUE}
swap("one", "two", "three")
```

Note that in this example we as passing three values,
not a single vector containing three values.
If we want a function to handle a [varying number of arguments](../glossary/#variable-arguments),
we represent the "extra" arguments with an ellipsis `...` (three dots),
which serves the same purpose as Python's `*args`:

```{r}
print_with_title <- function(title, ...) {
  title <- paste("==", title, "==\n")
  items <- paste(..., sep = "\n")
  cat(title)
  cat(items)
  cat("\n")
}
print_with_title("to-do", "Monday", "Tuesday", "Wednesday")
```

If we want to work with the extra arguments one by one,
we must first convert `...` to a list:

```{r}
add <- function(...) {
  result <- 0
  for (value in list(...)) {
    result <- result + value
  }
  result
}
add(1, 3, 5, 7)
```

A list in R is a vector that can contain values of many different types.
(The technical term for this is [heterogeneous](../glossary/#heterogeneous),
in contrast with the [homogeneous](../glossary/#homogeneous) nature of vectors' values.)

## How do I live without loops?

Modern Python encourages programmers to use [list comprehensions](../glossary/#list-comprehension) instead of loops,
e.g.,
to write:

```{python}
original = [3, 5, 7, 9]
doubled = [2 * x for x in original]
print(doubled)
```

instead of:

```{python}
doubled = []
for x in original:
  doubled.append(2 * x)
print(doubled)
```

If `original` is a NumPy array,
we can shorten this to:

```{python eval=FALSE}
doubled = 2 * original
```

As we have already seen,
R provides the same capability in the language itself:

```{r}
original <- c(3, 5, 7, 9)
doubled <- 2 * original
```

If we want to perform more complex calculations,
R encourages us to use a [functional](../glossary/#functional-programming) style of programming,
in which data is transformed by successive operations
much as reality itself is transformed by the will of the Great Old Ones.
We begin by defining a function that does what we want:

```{r}
double <- function(x) {
  2 * x
}
```

and then pass this function to `sapply` along with the data on which we want to operate:

```{r}
sapply(1:10, double)
```

This is clumsier than simply multiplying our values by two,
but it illustrates the way in which [higher-order functions](../glossary/#higher-order-function) work.
We can use other higher-order functions to perform other operations,
such as conditionals:

```{r}
values <- -5:5
ifelse(values > 0, values, NA)
```

R provides many other higher-order functions,
but their semantics are sometimes confusing.
We therefore encourage programmers to use the functions provided by the tidyverse instead;
these will be discussed in [the next lesson](../tidyverse/).

{% include links.md %}
