---
title: The Tidyverse
output:
  html_document:
    css: "languages.css"
---

```{r setup}
knitr::opts_chunk$set(engine.path = list(python = '/Users/gvwilson/anaconda3/bin/python'))
```

There is no point in becoming fluent in Akkadian if you do not then summon a Dweller Beneath
at the time of the new moon.
Similarly,
there is no point learning a language designed for data manipulation if you do not then bend data to your will.

## Reading Data

We begin by looking at the file `tidy/infant_hiv.csv`,
a tidied version of data on the percentage of infants born to women with HIV
who received an HIV test themselves within two months of birth.
The original data comes from the UNICEF site at <https://data.unicef.org/resources/dataset/hiv-aids-statistical-tables/>,
and this file contains:

```
country,year,estimate,hi,lo
AFG,2009,NA,NA,NA
AFG,2010,NA,NA,NA
...
AFG,2017,NA,NA,NA
AGO,2009,NA,NA,NA
AGO,2010,0.03,0.04,0.02
AGO,2011,0.05,0.07,0.04
AGO,2012,0.06,0.08,0.05
...
ZWE,2016,0.71,0.88,0.62
ZWE,2017,0.65,0.81,0.57
```

The actual file has many more rows and no ellipses.
It uses `NA` to show missing data rather than (for example) `-`, a space, or a blank,
and its values are interpreted as follows:

| Header | Datatype | Description |
|---|---|---|
| country | char | ISO3 country code of country reporting data |
| year | integer | year CE for which data reported |
| estimate | double/NA | estimated percentage of measurement |
| hi | double/NA | high end of range |
| lo | double/NA | low end of range |

```{python}
import pandas as pd

data = pd.read_csv('tidy/infant_hiv.csv')
print(data)
```

The equivalent in R is to load the tidyverse collection of libraries and then call the `read_csv` function.
We will go through this in stages, since each produces output.

```{r}
library(tidyverse)
```

Asking for the tidyverse gives us eight libraries (or *packages*).
One of those, dplyr, defines two functions that mask standard functions in R with the same names.
This is deliberate, and if we need the originals, we can get them with their fully-qualified names
`stats::filter` and `stats::lag`.
(Note that R uses `::` to get functions out of packages rather than Python's `.`.)

Once we have the tidyverse loaded,
reading the file looks remarkably like reading the file:

```{r}
data <- read_csv('tidy/infant_hiv.csv')
```

R's `read_csv` tells us more about what it has done than Pandas does.
In particular, it guesses at columns' data types based on the first thousand values,
and then tells us what types it has inferred.
(In a better universe,
people habitually use the first *two* rows of their spreadsheets for name *and units*,
but we do not live there.)

We can now look at what `read_csv` has produced.

```{r}
data
```

This is a *tibble*,
which is the tidyverse's enhanced version of R's `data.frame`.
It organizes data into named columns,
each having one value for each row.

## Data Inspection

We often have a quick look at the content of a table to remind ourselves what it contains.
Pandas does this using methods whose names are borrowed from the Unix shell's `head` and `tail` commands:

```{python}
print(data.head())
```
```{python}
print(data.tail())
```

R has similarly-named functions (not methods):

```{r}
head(data)
```
```{r}
tail(data)
```

Let's have a closer look at that last command's output:

```{r paged.print=FALSE}
tail(data)
```

Note that the row numbers printed by `tail` are relative to the output, not absolute to the table.
This is different from Pandas,
which retains the original row numbers.
(Notice also that R starts numbering from 1.)
What about overall information?

```{python}
print(data.info())
```

```{r}
summary(data)
```

Your display of R's summary may or may not wrap,
depending on how large a screen the older acolytes have allowed you.

## Indexing

A Pandas DataFrame is a collection of series (also called columns),
each containing the values of a single observed variable.
Columns in R tibbles are, not coincidentally, the same.

```{python}
print(data['estimate'])
```

We would get exactly the same output in Python with `data.estimate`,
i.e.,
with an attribute name rather than a string subscript.
The same tricks work in R:

```{r}
data['estimate']
```

However, `data$estimate` provides all the data

```{r}
data$estimate
```

Again, note that the boxed number on the left is the start index of that row.

What about single values?
Remembering to count from zero from Python and as humans do for R,
we have:

```{python}
print(data.estimate[11])
```
```{r}
data$estimate[12]
```

Ah—everything in R is a vector,
so we get a vector of one value as an output rather than a single value.

```{python error=TRUE}
print(len(data.estimate[11]))
```
```{r}
length(data$estimate[12])
```

And yes, ranges work:

```{python}
print(data.estimate[5:15])
```
```{r}
data$estimate[6:15]
```

Note that the upper bound is the same, because it's inclusive in R and exclusive in Python.
Note also that neither library prevents us from selecting a range of data that spans logical groups such as countries,
which is why selecting by row number is usually a sign of innocence, insouciance, or desperation.

We can select by column number as well.
Pandas uses the rather clumsy `object.iloc[rows, columns]`, with the usual `:` shortcut for "entire range":

```{python}
print(data.iloc[:, 0])
```

Since this is a column, it can be indexed:

```{python}
print(data.iloc[:, 0][0])
```

In R, a single index is interpreted as the column index:

```{r}
data[1]
```

But notice that the output is not a vector, but another tibble (i.e., an N-row, 1-column structure).
This means that adding another index does column-wise indexing on that tibble:

```{r}
data[1][1]
```

How then are we to get the first mention of Afghanistan?
The answer is to use double square brackets to strip away one level of structure:

```{r}
data[[1]]
```

This is now a plain old vector, so it can be indexed with single square brackets:

```{r}
data[[1]][1]
```

But that too is a vector, so it can of course be indexed as well (for some value of "of course"):

```{r}
data[[1]][1][1]
```

Thus,
`data[1][[1]]` produces a tibble,
then selects the first column vector from it,
so it still gives us a vector.
*This is not madness.*
It is merely…differently sane.

## Basic Statistics

What is the average estimate?
We start by grabbing that column for convenience:

```{python}
estimates = data.estimate
print(len(estimates))
```
```{python}
print(estimates.mean())
```

This translates almost directly to R:

```{r}
estimates <- data$estimate
length(estimates)
```
```{r}
mean(estimates)
```

It seems that the void is always there, waiting for us…
Let's fix this in R first:

```{r}
mean(estimates, na.rm=TRUE)
```

And then try to get the statistically correct behavior in Pandas:

```{python}
print(estimates.mean(skipna=False))
```

Many functions in R use `na.rm` to control whether `NA`s are removed or not.
(Remember, the `.` character is just another part of the name)
R's default behavior is to leave `NA`s in, and then to include them in aggregate computations.
Python's is to get rid of missing values early and work with what's left,
which makes translating code from one language to the next much more interesting than it might otherwise be.
But other than that, the statistics works the same way in Python:

```{python}
print(estimates.min())
```
```{python}
print(estimates.max())
```
```{python}
print(estimates.std())
```

Here are the equivalent computations in R:

```{r}
min(estimates, na.rm=TRUE)
```
```{r}
max(estimates, na.rm=TRUE)
```
```{r}
sd(estimates, na.rm=TRUE)
```

A good use of aggregation is to check the quality of the data.
For example,
we can ask if there are any records where some of the estimate, the low value, or the high value are missing,
but not all of them:

```{python}
print((data.hi.isnull() != data.lo.isnull()).any())
```

```{r}
any(is.na(data$hi) != is.na(data$lo))
```

## Filtering

By "filtering", we mean "selecting records by value".
As discussed [earlier](simple_things.md),
the simplest approach is to use a vector of logical values to keep only the values corresponding to `TRUE`.
In Python, this is:

```{python}
maximal = estimates[estimates >= 0.95]
print(len(maximal))
```

And in R:

```{r}
maximal <- estimates[estimates >= 0.95]
length(maximal)
```

The difference is unexpected.
Let's have a closer look at the result in Python:

```{python}
print(maximal)
```

And in R:

```{r}
maximal
```

It appears that R has kept the unknown values in order to highlight just how little we know—just how little
we *can* know.
More precisely,
wherever there was an `NA` in the original data
there is an `NA` in the logical vector
and hence an `NA` in the final vector.
Let us then turn to `which` to get a vector of indices at which a vector contains `TRUE`.
This function does not return indices for `FALSE` or `NA`:

```{r}
which(estimates >= 0.95)
```

And as a quick check:

```{r}
length(which(estimates >= 0.95))
```

So now we can index our vector with the result of the `which`:

```{r}
maximal <- estimates[which(estimates >= 0.95)]
maximal
```

But should we do this?
Those `NA`s are important information,
and should not be discarded so blithely.
What we should *really* be doing is using the tools the tidyverse provides
rather than clever indexing tricks.
These behave consistently across a wide scale of problems
and encourage use of patterns that make it easier for others to understand our programs.

## Tidy Code

The five basic data transformation operations in the tidyverse are:

- `filter`: choose observations (rows) by value(s)
- `arrange`: reorder rows
- `select`: choose variables (columns) by name
- `mutate`: derive new variables from existing ones
- `summarize`: combine many values to create a single new value

`filter(tibble, ...criteria...)` keeps rows that pass all of the specified criteria:

```{r}
filter(data, lo > 0.5)
```

Notice that the expression is `lo > 0.5` rather than `"lo" > 0.5`.
The latter expression returns the entire table
because the string `"lo"` is greater than the number 0.5 everywhere.
But wait:
how is it that `lo` can be used on its own?
It is the name of a column, but there is no variable called `lo`.

The answer is that R uses *lazy evaluation* of arguments.
Arguments aren't evaluated until they're needed,
so the function `filter` actually gets the expression `lo > 0.5`,
which allows it to check that there's a column called `lo` and then use it appropriately.
This is much tidier than `filter(data, data$lo > 0.5)` or `filter(data, "lo > 0.5")`,
and is *not* some kind of eldritch wizardry.
Many languages rely on lazy evaluation,
and when used circumspectly,
it allows us to produce code that is easier to read.

But we can do even better by using the pipe operator `%>%`,
which is about to become your new best friend:

```{r}
data %>% filter(lo > 0.5)
```

This may not seem like much of an improvement,
but neither does a Unix pipe consisting of `cat filename.txt | head`.
What about this?

```{r}
filter(data, (estimate != 0.95) & (lo > 0.5) & (hi <= (lo + 0.1)))
```

It uses the vectorized "and" operator `&` twice,
and parsing the condition takes a human being at least a few seconds.
Its tidyverse equivalent is:

```{r}
data %>% filter(estimate != 0.95) %>% filter(lo > 0.5) %>% filter(hi <= (lo + 0.1))
```

Breaking the condition into stages like this doesn't always make reading easier,
but it often helps development and testing.

Let's increase the band from 10% to 20%:

```{r}
data %>% filter(estimate != 0.95) %>% filter(lo > 0.5) %>% filter(hi <= (lo + 0.2))
```

And then order by `lo` in descending order,
breaking the line the way the [tidyverse style guide](http://style.tidyverse.org/) recommends:

```{r}
data %>%
  filter(estimate != 0.95) %>%
  filter(lo > 0.5) %>%
  filter(hi <= (lo + 0.2)) %>%
  arrange(desc(lo))
```

We can now add a `select` call to keep only the three columns we care about:

```{r}
data %>%
  filter(estimate != 0.95) %>%
  filter(lo > 0.5) %>%
  filter(hi <= (lo + 0.2)) %>%
  arrange(desc(lo)) %>%
  select(year, lo, hi)
```

Once again,
we are using the unquoted column names `year`, `lo`, and `hi`
and letting R's lazy evaluation take care of the details for us.

Rather than selecting these three columns,
we can select *out* the columns we're not interested in by negating their names.
This leaves the columns that are kept in their original order,
rather than putting `lo` before `hi`,
which won't matter if we later select by name,
but *will* if we ever want to select by position:

```{r}
data %>%
  filter(estimate != 0.95) %>%
  filter(lo > 0.5) %>%
  filter(hi <= (lo + 0.2)) %>%
  arrange(desc(lo)) %>%
  select(-country, -estimate)
```

Giddy with power,
we now add a column containing the difference between the low and high values.
This can be done using either `mutate`,
which adds new columns to the end of an existing tibble,
or with `transmute`,
which creates a new tibble containing only the columns we explicitly ask for.
Since we want to keep `hi` and `lo`,
we decide to use `mutate`:

```{r}
data %>%
  filter(estimate != 0.95) %>%
  filter(lo > 0.5) %>%
  filter(hi <= (lo + 0.2)) %>%
  arrange(desc(lo)) %>%
  select(-country, -estimate) %>%
  mutate(difference = hi - lo)
```

Does the difference between high and low estimates vary by year?
To answer that question,
we use `group_by` to put data into groups
and then `summarize` to summarize within groups.
Might as well get rid of the `arrange` and `select` calls in our pipeline at this point,
since we're not using them.
Oh,
and count how many records contributed to each aggregation using `n()`:

```{r}
data %>%
  filter(estimate != 0.95) %>%
  filter(lo > 0.5) %>%
  filter(hi <= (lo + 0.2)) %>%
  mutate(difference = hi - lo) %>%
  group_by(year) %>%
  summarize(n(), mean(year))
```

Let's do that again with more meaningful names for the final table's columns:

```{r}
data %>%
  filter(estimate != 0.95) %>%
  filter(lo > 0.5) %>%
  filter(hi <= (lo + 0.2)) %>%
  mutate(difference = hi - lo) %>%
  group_by(year) %>%
  summarize(count = n(), ave_diff = mean(year))
```

(We could also add a call to `rename`,
but for small tables like this,
setting column names on the fly is perfectly comprehensible.)

Now,
how might we do this with Pandas?
On approach is to use a single multi-part `.query` to select data
and store the result in a variable so that we can refer to the `hi` and `lo` columns twice
without repeating the filtering expression.
We then group by year and aggregate, again using strings for column names:

```{python}
data = pd.read_csv('tidy/infant_hiv.csv')
data = data.query('(estimate != 0.95) & (lo > 0.5) & (hi <= (lo + 0.2))')
data = data.assign(difference = (data.hi - data.lo))
grouped = data.groupby('year').agg({'difference' : {'ave_diff' : 'mean', 'count' : 'count'}})
print(grouped)
```

There are other ways to tackle this problem with Pandas,
but the tidyverse approach produces code that I find more readable.
