Our starting point is this:

1.  Each row in `percents` corresponds positionally to an ISO3 code in `countries`.
2.  Each group of three consecutive columns in `percents` has the estimate, high, and low values
    for a single year.
3.  The years themselves are not stored in `percents`,
    but we know from inspection that they start at 2009 and run without interruption to 2017.

Our strategy is to make a list of temporary tables:

1.  Take three columns at a time from `percents` to create a temporary tibble.
2.  Join `countries` to it.
3.  Create a column holding the year in each row and join that as well.

and then join those temporary tables row-wise to create our final tidy table.
(We will look later at how to do this without a loop,
but at this point in our journey it is probably safer to go step by step.)
Here is the addition to our script:

```{r output.lines=20}
first_year <- 2009
last_year <- 2017
num_years <- (last_year - first_year) + 1
chunks <- vector("list", num_years)
for (year in 1:num_years) {
  end <- year + 2
  temp <- select(percents, year:end)
  names(temp) <- c("estimate", "hi", "lo")
  temp$country <- countries
  temp$year <- rep((first_year + year) - 1, num_rows)
  temp <- select(temp, country, year, everything())
  chunks[[year]] <- temp
}
chunks
```

We start by giving names to our years;
if or when we decide to use this script for other data files,
we should extract the years from the data itself.
We then use `vector` to create the storage we are going to need to hold our temporary tables.
We could grow the list one item at a time,
but [allocating storage in advance](glossary.html#storage-allocation) is more efficient
and serves as a check on our logic:
if our loop doesn't run for the right number of iterations,
we will either overflow our list or have empty entries,
either of which should draw our attention.

Within the loop we figure out the bounds on the next three-column stripe,
select that,
and then give those three columns meaningful names.
This ensures that when we join all the sub-tables together,
the columns of the result will also be sensibly named.
Attaching the ISO3 country codes is as easy as assigning to `temp$country`,
and replicating the year for each row is easily done using the `rep` function.
We then reorder the columns to put country and year first
(the call to `everything` inside `select` selects all columns that aren't specifically selected),
and then we assign the temporary table to the appropriate slot in `chunks` using `[[..]]`.

As its name suggests,
`bind_rows` takes a list of tables and concatenates their rows in order.
Since we have taken care to give all of those tables the same column names,
no subsequent renaming is necessary.
We do,
however,
use `arrange` to order entries by country and year.

Now comes the payoff for all that hard work:

```{r}
tidy <- bind_rows(chunks)
tidy <- arrange(tidy, country, year)
tidy
```

What fresh hell is this?
Why do some rows have empty strings where country codes should be
and `NA`s for the three percentages?
Is our indexing off?
Have we somehow created one extra row for each year with nonsense values?

No.
It is not our tools that have failed us, or our reason, but our data.
("These parts are not fresh, Igor---I must have *fresh* parts to work with!")
Let us do this:

```{r}
raw <- read_csv("raw/infant_hiv.csv", skip = 2, na = c("-"))
missing <- raw %>%
  filter(is.na(Countries) | (Countries == "") | is.na(ISO3) | (ISO3 == "")) %>%
  select(Countries, ISO3)
missing
```

The lack of ISO3 country code for the region names doesn't bother us,
but Kosovo is definitely a problem.
[According to Wikipedia][wikipedia-iso3],
UNK is used for Kosovo residents whose travel documents were issued by the United Nations,
so we will fill that in with an ugly hack immediately after loading the data:

```{r}
raw <- read_csv("raw/infant_hiv.csv", skip = 2, na = c("-"))
raw$ISO3[raw$Countries == "Kosovo"] <- "UNK"
missing <- raw %>%
  filter(is.na(Countries) | (Countries == "") | is.na(ISO3) | (ISO3 == "")) %>%
  select(Countries, ISO3)
missing
```

All right.
Let's add that hack to our script,
then save the result to a file.
The whole thing is now 38 lines long:

```{r}
# Constants.
raw_filename <- "raw/infant_hiv.csv"
tidy_filename <- "tidy/infant_hiv.csv"
num_rows <- 192
first_year <- 2009
last_year <- 2017

# Get and clean percentages.
raw <- read_csv(raw_filename, skip = 2, na = c("-"))
raw$ISO3[raw$Countries == "Kosovo"] <- "UNK"
sliced <- slice(raw, 1:num_rows)
countries <- sliced$ISO3
body <- sliced %>%
  select(-ISO3, -Countries)
trimmed <- map_dfr(body, str_replace, pattern = ">?(\\d+)%", replacement = "\\1")
percents <- map_dfr(trimmed, function(col) as.numeric(col) / 100)

# Separate three-column chunks and add countries and years.
num_years <- (last_year - first_year) + 1
chunks <- vector("list", num_years)
for (year in 1:num_years) {
  end <- year + 2
  temp <- select(percents, year:end)
  names(temp) <- c("estimate", "hi", "lo")
  temp$country <- countries
  temp$year <- rep((first_year + year) - 1, num_rows)
  temp <- select(temp, country, year, everything())
  chunks[[year]] <- temp
}

# Combine chunks and order by country and year.
tidy <- bind_rows(chunks)
tidy <- arrange(tidy, country, year)

# Save.
write_csv(tidy, tidy_filename)
```

"It's alive!",
we exclaim,
but we can do better.
Let's start by using a pipeline for the code that extracts and formats the percentages:

```{r}
# Constants...

# Get and clean percentages.
raw <- read_csv(raw_filename, skip = 2, na = c("-"))
raw$ISO3[raw$Countries == "Kosovo"] <- "UNK"
sliced <- slice(raw, 1:num_rows)
countries <- sliced$ISO3
percents <- sliced %>%
  select(-ISO3, -Countries) %>%
  map_dfr(str_replace, pattern = ">?(\\d+)%", replacement = "\\1") %>%
  map_dfr(function(col) as.numeric(col) / 100)

# Separate three-column chunks and add countries and years...

# Combine chunks and order by country and year...

# Check...
```

The two changes are:

1.  We use a `%>%` pipe for the various transformations involved in creating percentages.
2.  We write the result to `temp.csv` so that we can compare it to the file created by our previous script.
    We should always do this sort of comparison when refactoring code in ways that isn't meant to change output;
    if the file is small enough to store in version control,
    we could overwrite it and use `git diff` or something similar to check whether it has changed.
    However,
    we would then have to trust ourselves to be careful enough not to accidentally commit changes,
    and frankly,
    we are no longer sure how trustworthy we are...

After checking that this has not changed the output,
we pipeline the computation in the loop:

```{r}
# Constans...

# Get and clean percentages...

# Separate three-column chunks and add countries and years.
num_years <- (last_year - first_year) + 1
chunks <- vector("list", num_years)
for (year in 1:num_years) {
  chunks[[year]] <- select(percents, year:(year + 2)) %>%
    rename(estimate = 1, hi = 2, lo = 3) %>%
    mutate(country = countries,
           year = rep((first_year + year) - 1, num_rows)) %>%
    select(country, year, everything())
}

# Combine chunks and order by country and year.
tidy <- bind_rows(chunks) %>%
  arrange(country, year)
```

We have introduced a call to `rename` here to give the columns of each sub-table the right names,
and used `mutate` instead of assigning to named columns one by one.
The lack of intermediate variables may make the code harder to debug using print statements,
but certainly makes this incantation easier to read aloud.

So we run it and inspect the output and it's the same as what we had
and we're about to commit to version control
when we decide to double check against the original data and guess what?
The values for Argentina are wrong.
In fact,
the values for most countries and years are wrong:
only the ones in the first three columns are right.
The problem,
it turns out,
is that our loop index `year` is going up in ones,
while each year's data is three columns wide.
Here's the final, *final*, __*final*__ version:

```{r}
library(tidyverse)

# Constants.
raw_filename <- "raw/infant_hiv.csv"
tidy_filename <- "tidy/infant_hiv.csv"
first_year <- 2009
last_year <- 2017
num_rows <- 192

# Get and clean percentages.
raw <- read_csv(raw_filename, skip = 2, na = c("-"))
raw$ISO3[raw$Countries == "Kosovo"] <- "UNK"
sliced <- slice(raw, 1:num_rows)
countries <- sliced$ISO3
percents <- sliced %>%
  select(-ISO3, -Countries) %>%
  map_dfr(str_replace, pattern = ">?(\\d+)%", replacement = "\\1") %>%
  map_dfr(function(col) as.numeric(col) / 100)

# Separate three-column chunks and add countries and years.
num_years <- (last_year - first_year) + 1
chunks <- vector("list", num_years)
for (year in 1:num_years) {
  start = 3 * (year - 1) + 1
  chunks[[year]] <- select(percents, start:(start + 2)) %>%
    rename(estimate = 1, hi = 2, lo = 3) %>%
    mutate(country = countries,
           year = rep((first_year + year) - 1, num_rows)) %>%
    select(country, year, everything())
}

# Combine chunks and order by country and year.
tidy <- bind_rows(chunks) %>%
  arrange(country, year)

# Check.
write_csv(tidy, tidy_filename)
```
