---
title: "Reticulate"
output: md_document
questions:
  - "How can I use Python and R together?"
objectives:
  - "Use `reticulate` to share data between R and Python."
  - "Use `reticulate` to call Python functions from R code and vice versa."
  - "Run Python scripts directly from R programs."
keypoints:
  - "The `reticulate` library allows R programs to access data in Python programs and vice versa."
  - "Use `py.whatever` to access a top-level Python variable from R."
  - "Use `r.whatever` to access a top-level R definition from Python."
  - "R is always indexed from 1 (even in Python) and Python is always indexed from 0 (even in R)."
  - "Numbers in R are floating point by default, so use a trailing 'L' to force a value to be an integer."
  - "A Python script run from an R session believes it is the main script, i.e., `__name__` is `'__main__'` inside the Python script."
---

```{r setup, include=FALSE}
source("common.R")
library(tidyverse)
```

As the previous lessons have shown,
you can do a lot with R,
but sometimes you might feel a cold, serpentine tug on your soul
pulling you back to Python.
You can put Python code in RMarkdown documents:

```{python}
print("Hello R")
```

but how can those chunks interact with your R and vice versa?
The answer is a package called [reticulate][reticulate] that provides two-way communication between Python and R.
To use it,
run `install.packages("reticulate")`.
By default,
it uses the system-default Python:

```{r}
Sys.which("python")
```

but you can configure it to use different versions,
or to use `virtualenv` or a Conda environment---see [the document][reticulate-configure] for details.

## How can I access data across languages?

The most common way to use reticulate is to do some calculations in Python and then use the results in R
or vice versa.
To show how this works,
let's read our infant HIV data into a Pandas data frame:

```{python}
import pandas
data = pandas.read_csv('tidy/infant_hiv.csv')
print(data.head())
```

All of our Python variables are available in our R session as part of the `py` object,
so `py$data` is our data frame inside a chunk of R code:

```{r}
library(reticulate)
head(py$data)
```

reticulate handles type conversions automatically,
though there are a few tricky cases:
for example,
the number `9` is a float in R,
so if you want an integer in Python,
you have to add the trailing `L` (for "long") and write it `9L`.

On the other hand,
reticulate translates between 0-based and 1-based indexing.
Suppose we create a character vector in R:

```{r}
elements = c('hydrogen', 'helium', 'lithium', 'beryllium')
```

Hydrogen is in position 1 in R:

```{r}
elements[1]
```

but position 0 in Python:

```{python}
print(r.elements[0])
```

Note our use of the object `r` in our Python code:
just `py$whatever` gives us access to Python objects in R,
`r.whatever` gives us access to R objects in Python.

## How can I call functions across languages?

We don't have to run Python code,
store values in a variable,
and then access that variable from R:
we can call the Python directly (or vice versa).
For example,
we can use Python's random number generator in R as follows:

```{r}
pyrand <- import("random")
pyrand$gauss(0, 1)
```

(There's no reason to do this---R's random number generator is just as strong---but it illustrates the point.)

We can also source Python scripts.
For example,
suppose that `countries.py` contains this function:

```{python code=readLines('countries.py'), eval=FALSE}
```

We can run that script using `source_python`:

```{r}
source_python('countries.py')
```

There is no output because all the script did was define a function.
By default,
that function and all other top-level variables defined in the script are now available in R:

```{r}
get_countries('tidy/infant_hiv.csv')
```

There is one small pothole in this.
When the script is run,
the special Python variable `__name__` is set to `'__main__'"'`,
i.e.,
the script thinks it is being called from the command line.
If it includes a conditional block to handle command-line arguments like this:

```{python eval=FALSE}
if __name__ == '__main__':
    input_file, output_files = sys.argv[1], sys.argv[2:]
    main(input_file, output_files)
```

then that block will be executed,
but will fail because `sys.argv` won't include anything.

{% include links.md %}